Poker Chips, Earthquakes, and Computation

### 2014-10-13 finaldraft.tex
added commas and periods after equations
typo fixes

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\begin{equation*}
\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2 
\end{equation*}

@@

\begin{lemma}
\label{spreadfunction}
Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb \neq b$. Then
\begin{equation*}
\Psi(E_sb) > \Psi(b) 
\end{equation*}
so the earthquake strictly increases the spread of the board.
\end{lemma}
\begin{proof}
The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.

@@

Omitting the algebraic details, we find that summing up the changes in the spread function gives
\begin{equation*}
 \Psi(E_sb) - \Psi(b) = 4
 \end{equation*}
which completes the proof.
\end{proof}

@@

For ease of notation we also introduce an indicator function
\begin{definition}
We define the indicator function $[[\cdot]]$ as 
\begin{equation*}
[[A]] = \begin{cases} 1 \text{ if A is true} \\ 0 \text{ if A is false} \end{cases}
\end{equation*}
\end{definition}

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$. If $E$ is a valid earthquake, then 
\begin{equation*}
E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]]
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

@@

\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2 E_1 b = E_1 E_2 b
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

@@

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in Lemma~\ref{swappinglemma}. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
E_m, E_{m+1}, \ldots, E_{j-1}, E_j
\end{equation*}
where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
E_{j}, E_{m+1}, \ldots,  E_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma~\ref{swappinglemma} we know that the two sequences 
\begin{align*}
E_{m}, E_{j} \\
E_{j}, E_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction we used in our first step of this proof we have that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board. 
\end{proof}

@@

Thus we have transformed our starting sequence into a sequence that is completely determined by the state of the initial board, noting that the final state of the board is still the same. We can do this for any other sequence that stabilizes the board and get the same sequence. This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes it never adds or removes them). 
\end{proof}

@@

We now define a coordinate system so that our method of proof can mirror that of the standard board case.
\begin{definition} We define a coordinate system for our joined $d$-trees, with the following rules:
\begin{enumerate}
	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq (d-1)^{|a|}$
	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq (d-1)^{|a-1|}$
\end{enumerate}
We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
\end{definition}
\begin{equation*}
G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}
\end{equation*}

We now formalize the board of poker chips. Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
\begin{definition}
We define a $d$-board to be a function from the set of valid coordinates to the nonnegative integers, that is 
\begin{equation*}
b:  G_d \rightarrow \mathbb{Z}_{\geq 0}
\end{equation*}
\end{definition}

@@

We once again define a spread function 
\begin{definition}
$\Psi$ is a function from the finite $d$-boards and is defined as
\begin{equation*}
\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2
\end{equation*}
Note that this is well defined for boards which contain a finite number of chips. 
\end{definition} 
We can once again think of this spread function as the sum of the number of chips on every node weighted by the taxi-cab distance from the node to the origin.

We again demonstrate that nontrivial earthquakes increase the value of the spread function, using a nearly identical proof.
\begin{lemma} 
Suppose an earthquake hits node $s$ of a finite $d$-board $b$ producing the board $E_s b$. Then
\begin{equation*}
\Psi( E_s b) > \Psi (b)
\end{equation*}
\end{lemma}
\begin{proof}
Once again the chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus the contribution to $\Psi$ is the same at all nodes except for those $d+1$.

@@

Omitting the algebraic details we find that 
\begin{align*}
\Psi( E_s b) - \Psi(b) > d >0
\end{align*}
which completes the proof.
\end{proof}

@@

In this section, we show how we can compute introduce the ideas of Boolean functions and Boolean circuits and show how we can compute these use the poker chip board and the earthquake operator.

@@

We now define a simple model of computation. We assume a finite board and specify some nodes as inputs nodes and some as outputs. If an input bit is to be set to 1, then we place 4 chips on the input node. Otherwise we do not place any chips on it. We then apply global earthquakes to the board until it stabilizes. If an output node ever has chips redistributed from it by an earthquake, then we set the output bit to 1; otherwise, the output bit is set to 0.

A simple example of this is a board that computes the identity function of one input. We can compute 
\begin{equation*}
f(x) = x
\end{equation*}
Our board has no chips on it, other than perhaps at the input node. We set the output node to be the same as the input node. If $x = 0$, then the input node does not have any chips on it and the board is stable. An earthquake never occurred on the output node and so the output is 0. If $x = 1$, then we place 4 chips on the input node and in order for the board to stabilize an earthquake must effect it. Because the input and output node are the same, this means an earthquake occurred on the output node and so the output is 1.

Using this simple model, there are a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).

@@

\begin{theorem}
There does not exists a board which computes $f(x) = \neg x$.
\end{theorem}

@@

Let $A(0)$ be the set of nodes that are targeted by earthquakes when $i = 0$ and $A(1)$ be the set of nodes that are targeted by earthquakes when $i=1$.  We claim that $A(0) \subset A(1)$. This is true as when we set the input square to 1 we only increase the locations that can have chips redistributed from them by earthquakes. However the output square is not in $A(1)$ by the definition of our function, which produces a contradiction. Therefore our current model of computation cannot have a board that computes NOT. 
\end{proof}

@@

\noindent Another earthquake\\
hits the board, and then one more\\
Who cares which was first! \\

\noindent Now the aftershock\\
is computing some functions,\\
but just basic ones.
\end{document}

>>>>

To define the spread function, we fix one coordinate on a board $b$ as the origin $(0,0)$, and we define the \emph{taxicab distance} $d(s)$ of a node $s$ on the board to be $|s_x| + |s_y|$, where $s_x$ is the $x$ coordinate of node $s$, and $s_y$ is the $y$ coordinate of node $s$. Then the spread function is
\begin{equation*}
\Psi(b) = \sum_{\text{all nodes }s}b(s)d(s)^2.
\end{equation*}

@@

\begin{lemma}
\label{spreadfunction}
Suppose an earthquake hits node $s$ of a finite board $b$, producing the board $E_sb \neq b$. Then
\begin{equation*}
\Psi(E_sb) > \Psi(b),
\end{equation*}
so the earthquake strictly increases the spread of the board.
\end{lemma}
\begin{proof}
The chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus, the contribution to $\Psi$ is the same at all nodes except for those 5.

@@

Omitting the algebraic details, we find that summing up the changes in the spread function gives
\begin{equation*}
 \Psi(E_sb) - \Psi(b) = 4,
 \end{equation*}
which completes the proof.
\end{proof}

@@

For ease of notation we also introduce an indicator function
\begin{definition}
We define the indicator function $[[\cdot]]$ as 
\begin{equation*}
[[A]] = \begin{cases} 1 \text{ if A is true} \\ 0 \text{ if A is false} \end{cases}.
\end{equation*}
\end{definition}

Recalling that $n(s)$ refers to the set of neighbors of the square $s$ we note that if an earthquake $E$ is valid then we can write it in terms of this indicator function.
\begin{lemma} 
\label{earthquakeredefine}
Consider a board $b$. If $E$ is a valid earthquake, then 
\begin{equation*}
E_s b(x) = b(x) + 1[[ x \in n(s)]] - 4[[x =s]].
\end{equation*}
\end{lemma}
\begin{proof}
This follows directly from the definition of an earthquake and a valid earthquake. 
\end{proof}

@@

\begin{lemma}
\label{swappinglemma}
Consider a board $b$ and two earthquakes $E_1$ and $E_2$, affecting nodes $s_1$ and $s_2$ respectively. If $E_1, E_2$ is a valid sequence and $E_2$ is a valid earthquake when applied before $E_1$ then $E_2, E_1$ is also a valid sequence of earthquakes and 
\begin{equation*}
E_2E_1 b = E_1E_2 b.
\end{equation*}
\end{lemma}
\begin{proof}
We will first show that $E_2, E_1$ is a valid sequence and then use this to show that the order in which the earthquakes occur does not matter.

@@

We can show this by inducting on the length of the sequence $l=j-m+1$. The base case of $l=2$ is shown in Lemma~\ref{swappinglemma}. Let us assume the result holds for sequences of length $l=p$ and suppose our sequence has length $l=p+1$. So our sequence is 
\begin{equation*}
E_m, E_{m+1}, \ldots, E_{j-1}, E_j,
\end{equation*}
where $p+1=j-m+1$. Then if we set $b_{m} = E_{m} (b_{m-1})$ by applying our inductive step we have that the two sequences 
\begin{align*}
E_{m+1}, \ldots, E_{j-1}, E_{j} \\ 
E_{j}, E_{m+1}, \ldots,  E_{j-1} 
\end{align*}
are both valid and produce the same final board configuration. Now if we consider the board state $b_{m-1}$ and once again apply Lemma~\ref{swappinglemma} we know that the two sequences 
\begin{align*}
E_{m}, E_{j} \\
E_{j}, E_{m}
\end{align*}
are both valid and produce the same final board configuration. If we apply the same reduction we used in our first step of this proof we have that $E_j, E_m, E_{m+1}, \ldots, E_{j-1}$ and $E_m, E_{m+1}, \ldots, E_{j-1}, E_j$ both applied to $b_{m-1}$ are valid and produce the same board. 
\end{proof}

@@

Thus we have transformed our starting sequence into a sequence that is completely determined by the state of the initial board, noting that the final state of the board is still the same. We can do this for any other sequence that stabilizes the board and get the same sequence. This shows not only that the order in which the earthquakes occur does not change the final state of the board but also that a given square is targeted by the same number of valid earthquakes regardless of the order of the earthquakes (we know this because our process only shifts the order of valid earthquakes; it never adds or removes them). 
\end{proof}

@@

We now define a coordinate system so that our method of proof can mirror that of the standard board case.
\begin{definition} We define a coordinate system for our joined $d$-trees, with the following rules:
\begin{enumerate}
	\item $(0,0)$ and $(1,0)$ are valid coordinates and refer to the two parent nodes.
	\item $(a,b)$ is a valid coordinate for $a<0$ if $0<b \leq (d-1)^{|a|}$
	\item $(a,b)$ is a valid coordinate for $a>1$ if $0<b \leq (d-1)^{|a-1|}$
\end{enumerate}
We define the set of all valid coordinates $G_d$ for a joined $d$-tree as 
\end{definition}
\begin{equation*}
G_d:= \{ (a,b) \in \mathbb{Z} \times \mathbb{Z} ; (a,b) \text{ obey the rules above}\}.
\end{equation*}

We now formalize the board of poker chips. Note that this definition only differs from our initial definition in that the input value is a coordinate of a node in a joined $d$-tree rather than a coordinate of the cartesian grid.
\begin{definition}
We define a $d$-board to be a function from the set of valid coordinates to the nonnegative integers, that is 
\begin{equation*}
b:  G_d \rightarrow \mathbb{Z}_{\geq 0}.
\end{equation*}
\end{definition}

@@

We once again define a spread function 
\begin{definition}
$\Psi$ is a function from the finite $d$-boards and is defined as
\begin{equation*}
\Psi(b) = \sum_{(i,j) \in G_d} b(i,j) |i|^2.
\end{equation*}
Note that this is well defined for boards which contain a finite number of chips. 
\end{definition} 
We can once again think of this spread function as the sum of the number of chips on every node weighted by the taxi-cab distance from the node to the origin.

We again demonstrate that nontrivial earthquakes increase the value of the spread function, using a nearly identical proof.
\begin{lemma} 
Suppose an earthquake hits node $s$ of a finite $d$-board $b$ producing the board $E_s b$. Then
\begin{equation*}
\Psi( E_s b) > \Psi (b).
\end{equation*}
\end{lemma}
\begin{proof}
Once again the chip stacks of the two boards are the same at all nodes except at $s$ and $n(s)$, the nodes that neighbor $s$. Thus the contribution to $\Psi$ is the same at all nodes except for those $d+1$.

@@

Omitting the algebraic details we find that 
\begin{align*}
\Psi( E_s b) - \Psi(b) > d >0,
\end{align*}
which completes the proof.
\end{proof}

@@

In this section, we show how we can compute Boolean functions using the poker chip problem setup. In particular, we use the fact that Boolean functions can be represented as Boolean circuits, and we show how to construct elements of Boolean circuits using the poker chip board and earthquake operator.

@@

We now define a simple model of computation. We assume a finite board and specify some nodes as inputs nodes and some as outputs. If an input bit is to be set to 1, then we place 4 chips on the input node. Otherwise we do not place any chips on it. We then apply global earthquakes to the board until it stabilizes. If an output node ever has chips redistributed to it by an earthquake, then we set the output bit to 1; otherwise, the output bit is set to 0.

A simple example of this is a board that computes the identity function of one input. We can compute 
\begin{equation*}
f(x) = x.
\end{equation*}
Our board has no chips on it, other than perhaps at the input node. We set the output node to be the same as the input node. If $x = 0$, then the input node does not have any chips on it and the board is stable. An earthquake never occurred on the output node and so the output is 0. If $x = 1$, then we place 4 chips on the input node and in order for the board to stabilize an earthquake must effect it. Because the input and output node are the same, this means an earthquake occurred on the output node and so the output is 1.

Using this simple model, there is a surprising number of things we can do. We can make wires, AND gates and OR gates (see Figure~\ref{fig:simplegates}).

@@

\begin{theorem}
There does not exist a board which computes $f(x) = \neg x$.
\end{theorem}

@@

Let $A(0)$ be the set of nodes that are targeted by earthquakes when $i = 0$ and $A(1)$ be the set of nodes that are targeted by earthquakes when $i=1$.  We claim that $A(0) \subset A(1)$. This is true as when we set the input square to 1, we only increase the locations that can have chips redistributed to them by earthquakes. However the output square is not in $A(1)$ by the definition of our function, which produces a contradiction. Therefore our current model of computation cannot have a board that computes NOT. 
\end{proof}

@@

\noindent Another earthquake\\
hits the board, and then one more.\\
Who cares which was first! \\

\noindent Now the aftershock\\
is computing some functions\\
---but just basic ones!
\end{document}
